---
title: "Class 12 HW"
author: "Charlie Rezanka (A15837296)"
format: html
---

# Proportion of G/G in data

Downloaded a CSV file from ensemble (https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39905111-40015112;v=rs8067378;vdb=variation;vf=105535077#373531_tablePanel)

```{r}
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv", row.names = 1)

head(mxl)
```


```{r}
table(mxl$Genotype..forward.strand.)
```

```{r}
(table(mxl$Genotype..forward.strand.) / nrow(mxl))*100
```

Lets look at a different population, such as Great Britin

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv", row.names =1)

head(gbr)
```

Find the portion of G/G 

```{r}
(table(gbr$Genotype..forward.strand.)/ nrow(gbr))*100
```

The haplotype most associated with childhood asthma is more common in the GB population than MXL


## Population Scale Analysis Homework

```{r}
exp <- read.table("rs8067378_ENSG00000172057.6.txt", row.names = 1)
head(exp)
```

> Q13. Read this file into R and determine the sample size for each genotype and their corresponding median expression levels for each of these genotypes.

```{r}
nrow(exp)
```


```{r}
table(exp$geno)
```

```{r}
medians <- aggregate(exp ~ geno, data = exp, FUN = median)
head(medians)
```


Out of a total of 462 samples, the A/A genotype has a sample size of 108, A/G of 233, and G/G of 121. Median expression for A/A is 31.2, for A/G is 25.1, and G/G is 20.1.



> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

```{r}
library(ggplot2)
ggplot(exp) + aes(geno, exp, fill=geno) + 
  geom_boxplot()
```

The presence of the G allele in either heterozygous or homozygous genotypes will result in a drop in ORMDL3 expression, with the homozygous G/G genotype having the lowest expression. 





