---
title: "Week 3, R Functions"
author: "Charlie"
date: "2023-04-22"
output: html_document
---

This week we're intorducing R functions and how to write our own.

> **Q1.** Write a function grade() to determine an overall grade from a vector of student homework assignment scores dropping the lowest single score. If a student misses a homework (i.e. has an NA value) this can be used as a score to be potentially dropped.

```{r}
# Example input vectors to start with
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)

```

Guidelines from class
- write a working code snippet to solve a simple problem

```{r}
#Straight forward  mean
student1 <- c(100, 100, 100, 100, 100, 100, 100, 90)
mean(student1)
```

However, we need to identify and drop the lowest score for each student. 

```{r}
#Which element of the student1 vector is the lowest?
which.min(student1)
```

If I want to drop (exclude) the lowest score from my mean calculation:

```{r}
#This will return everything BUT the 8th element of my vector
student1[-8]
```

Now we can use the answer from which.min to find the answer to our question.

```{r}
#Here is our first snippet of code. 
mean(student1[-which.min(student1)])
```


For Students 2 and 3 we will have to replace all NAs with 0


Student 2:

```{r}
student2 <- c(100, NA, 90, 90, 90, 90, 97, 80)
x <- student2
is.na(x)
which( is.na(x))
```

We've now identified the NA values we want to mask and replace NA with 0.

```{r}
x[is.na(x)] <- 0
x
mean(x)
```

Now we just drop the lowest score. 

```{r}
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

Student 3: 

```{r}
student3 <- c(90, NA, NA, NA, NA, NA, NA, NA)
x <- student3
x[is.na(x)] <- 0
mean(x[-which.min(x)])
```

Now we're nearly there with our snippet! 

##Now take the snippet and turn it into a function

Every function has 3 parts:
- A name (in our case `grade()`)
- Input arguments, a vector of student scores
- The body -- our snippet


I will do this by selecting `Code then Extract Function` in Rstudio
```{r}
grades <- function(x) {
  x[is.na(x)] <- 0
  mean(x[-which.min(x)])
}
```


```{r}
grades(student1)
grades(student2)
grades(student3)
```

The code works! We just now need to add our comments to the function so that we can come back to this and understand it. 

```{r}
#' Calculate the average scroe for a vector of student scores, dropping the lowest value.
#' Missing values will be treated as 0
#'
#' @param x A numeric vector of student scores. 
#'
#' @return Average score
#' @export
#'
#' @examples
#' student <- c(100, NA, 90, 97)
#' grade(student)
#' 
#' 
grades <- function(x) {
  #treat missing values as 0
  x[is.na(x)] <- 0
  #exclude lowest value from mean
  mean(x[-which.min(x)])}
```



> **Q2** Using your grade() function and the supplied gradebook, Who is the top scoring student overall in the gradebook?

```{r}
url <-  "https://tinyurl.com/gradeinput"
gradebook <- read.csv(url, row.names = 1)

```


```{r}
apply(gradebook, 1, grades)
```

Can calculate highest perfomrer using the apply() function
```{r}
results <- apply(gradebook, 1, grades)
results
```

```{r}
which.max(results)
```



**The highest performing student was student-18**



> **Q3.** Which homeowrk was the toughest?? 


```{r}
avg.scores <- apply(gradebook, 2, mean, na.rm=TRUE)
avg.scores
which.min(avg.scores)


```

```{r}
med.scores <- apply(gradebook, 2, median, na.rm=TRUE)
med.scores
which.min(med.scores)
```

Which is better? 

```{r}
boxplot(gradebook)
```


Due to variation in the data, the median is a more reliable metric to use when answering this question. 

**The toughest homeowrk assignment was HW2**



> **Q4.** From your analysis of the gradebook, which homework was most predictive of overall score? 


Are the final results (avg score distribtuion) correlated with the results for individual homeworks? 
```{r}
masked.gradebook <- gradebook
masked.gradebook[is.na(masked.gradebook)] <- 0 
masked.gradebook

```

```{r}
cor(results, gradebook$hw1)
```

```{r}
apply(masked.gradebook, 2, cor, x=results)
```

**HW5 is the best predictor of overall class grades**




> **Q5.** Make sure you save your Quarto document and can click the “Render” (or Rmarkdown”Knit”) button to generate a PDF foramt report without errors. Finally, submit your PDF to gradescope. 

Knit the document to produce a PDF *(or HTML that you'll convert to PDF later if your computer is trash like mine)*







