---
title: "Class 11 - Mini Halloween Project"
author: "Charlie Rezanka (A15837296)"
format: html
---


In today's class, we'll use some sample data about halloween candy preferances to get a greater feeling for how PCA and other methods work.



```{r}
candy <- read.csv("candy-data.txt", row.names = 1)
head(candy)

```
#the winpercent catagory is how frequently people indicate a preference for a candy
#pricepercent analyzes the cost of a candy relative to others



> Q1. How many candy types are in this data?

```{r}
nrow(candy)
```



> Q2. How many different types fof fruity candy are in this data? 

```{r}
table(candy$fruity)
```

What are these fruity candies in question?

we can use ==
```{r}
row.names(candy[candy$fruity == 1, ])
```



## How often does my favorite candy win in these matchups?

```{r}
candy["Twix", "winpercent"]
```



> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?


```{r}
candy["Milky Way", "winpercent"]
```


> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", "winpercent"]
```


> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", "winpercent"]
```

```{r}
install.packages("skimr", repos = "http://cran.us.r-project.org")
library("skimr")
skim(candy)
```


> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The wineprcent variable is on a completely different scale to every other variable.



> Q7. What do you think a zero and one represent for the candy$chocolate column?

Whether or not a candy contains chocolate (yes = 1)


> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy) +
  aes(winpercent) +
  geom_histogram(bins=10, fill="BLUE")
```

```{r}
help(ggplot)
```

> Q9. Is the distribution of winpercent values symmetrical?

The distribution of winpercent values is baised towards the 30-50% region, it is not symetrical. 

> Q10. Is the center of the distribution above or below 50%?

```{r}
median(candy$winpercent)
```


> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
#Filter your data first to be just chocolate
chocolate.candy <- candy[as.logical(candy$chocolate),]

#Get the winpercent values
chocolate.winpercent <- chocolate.candy$winpercent

#calculate the mean
mean(chocolate.winpercent)
sd(chocolate.winpercent)
```

```{r}
fruity.candy <- candy[as.logical(candy$fruity),]
fruity.winpercent <- fruity.candy$winpercent
mean(fruity.winpercent)
sd(fruity.winpercent)
```

Chocolate candy is more popular than fruity candy



> Q12. Is this difference statistically significant?


```{r}
t.test(chocolate.winpercent, fruity.winpercent)
```

They are not significantly different



## OVerall Candy Ratings

There is a base R function called `sort()` for sorting vectors of input. 

```{r}
x <- c (5,2,10)

#sort(x, decreasing = FALSE) by default (orders by increasing)
sort (x)
```

The buddy function to `sort()` that is often even more useful is called `order()`. It returns the "indices" of the input that would result in it being sorted (the numbered vector items placed in decreasing order)

```{r}
order(x)
x[order(x)]
```

> Q13. What are the five least liked candy types in this set?

I can order by winpercent 
```{r}
ord <- order(candy$winpercent)
head(candy[ord,],5)
```




> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
ord <- order(candy$winpercent, decreasing=TRUE)
head(candy[ord,],5)
```



> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()
```


> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col()
```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy), winpercent)) +
  geom_col(fill=my_cols)
```






> Q17. What is the worst ranked chocolate candy?

Sixlets


> Q18. What is the best ranked fruity candy?

Starbursts



## Taking a look at pricepercent

> Q. What's the best candy for the least amout of money?

```{r}
install.packages("ggrepel", repos = "http://cran.us.r-project.org")
```


```{r}
library(ggrepel)
my_cols[as.logical(candy$fruity)] = "red"

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
   geom_text()
```


To negate overlapping labels, I can use the ggrepel package to redesign my labels

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) +
   geom_text_repel(max.overlaps=8, col=my_cols, size=3)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Recee's miniatures


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nip (least popular), Ring pop, Nestle smarties, Mr good bar, whoppers




## Exploring the correlation structure

pearson correlation goes between -1 and +1 with 0 indicating no correlation, and values close to one being very highly (ani) correlated. 

```{r}
install.packages("corrplot", repos = "http://cran.us.r-project.org")
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruity candies


> Q23. Similarly, what two variables are most positively correlated?

Chocolate & bar with pricepercent and winpercent




## Principal Component Analysis

The base R function for PCA is `prcomp()` and we can set "scale=TRUE/FALSE". 

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

The main result of PCA (i.e. the new PC plot) is contained in `pca$x()`


```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) +
  aes(PC1, PC2, label=rownames(pc)) +
  geom_point(col=my_cols) +
  geom_text_repel(max.overlaps = 7, size=3)
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus were all picked up with a positive correlation in the PC analysis. This makes sense as all fruity candies were very closely grouped on our PCA plot, and many of the popular fruity candies I can think of come in packages of many small candies (ex. sour patch kids & starbursts)










